A Review of Classic Edge Detectors

### 2014-11-11 old_versions/edge-detection-review-v6.tex
second round of third revision, improving references

%-------------------------------------------------------------------------------
% Edge detectors review v5
% by Haldo Spontón & Juan Cardelino
% IPOL 2012
%-------------------------------------------------------------------------------
\documentclass{ipol}
\ipolSetYear{2012}
\ipolSetDOI{10.XXXX/ipol.2012.sc-red}

\usepackage{hyperref,verbatim,graphicx,amsmath,amssymb,amssymb,dsfont}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{sidecap,subfig,array}
\usepackage[table]{xcolor}
\usepackage{multicol}
\usepackage{listings}
\usepackage{color}
\newtheorem{theorem}{Theorem}
\numberwithin{equation}{section}
\numberwithin{table}{section}
\numberwithin{figure}{section}
\begin{document}
\bibliographystyle{ieeetr}
\lstset{
	backgroundcolor=\color[rgb]{0.94,0.94,0.94},
    tabsize=4,
    inputencoding=utf8x,
         extendedchars=\true,
         language=C,
         basicstyle=\scriptsize,
                 showtabs=false,
        showspaces=false,
        showstringspaces=false,
        identifierstyle=\ttfamily,
        keywordstyle=\color[rgb]{0,0,1},
        commentstyle=\color[rgb]{0.133,0.545,0.133},
        stringstyle=\color[rgb]{0.627,0.126,0.941},
}

%-------------------------------------------------------------------------------
\title{Review of Edge Detectors}
\author{Haldo Spont\'on,\\
        Juan Cardelino}
\date{}
\ipolMaketitle

%-------------------------------------------------------------------------------
\begin{ipolAbstract}
In this paper we study some of the classic alternatives for edge detection in digital images. The main idea 
behind edge detection is to find where there are abrupt changes in the intensity of an image. 
As a first approach, the first derivative is used to find the changes of intensity in several algorithms 
such as Sobel, Prewitt and Roberts. The second derivative is also used, for example in algorithms 
like Marr-Hildreth and Haralick.\\
We compare the results obtained from a qualitative point of view (perceptual) and from a quantitative 
point of view (number of operations, execution time), considering different forms to convolve an 
image with a kernel (step required in some of the algorithms).\\
\end{ipolAbstract}

%-------------------------------------------------------------------------------
\begin{ipolCode}
The C implementation of the algorithms implemented in this work (beta version) can be
downloaded from \href{http://iie.fing.edu.uy/~haldos/downloads/edge_detectors_v0.1.tar.gz}{here}.
You can also follow this code project in this GIT repository: 
\begin{center}
\url{git://github.com/haldos/edges.git}
\end{center}
\end{ipolCode}

%-------------------------------------------------------------------------------
%\begin{ipolSupp}
%\end{ipolSupp}

%-------------------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

@@

In the 70's, edge detection methods were based on using small operators 
(such as Sobel masks), attempting to compute an approximation of the
first derivative of the image \cite{Gonzalez2007Digital}. In section ~\ref{sec:first} we study the 
performance of such algorithms, as an introduction to a more sophisticated 
analysis into the edge finding process.\\

In 1980 Marr and Hildreth \cite{AIM-518} argued that intensity changes are not independent 
of image scale, so the edge detection requires the use of different size 
operators. They also argued that a sudden intensity change will be seen 
as a peak (or trough) in the first derivative or, equivalently, as a zero 
crossing in the second derivative. The bulk of this work focuses on the 
study of this algorithm, presented in Section ~\ref{sec:second}. In this section 
we also present the Haralick algorithm \cite{bb20239}, which is also based on the second derivative.
This algorithm has the particularity of proposing a model to locally approximate 
the image around a point. Then, with this model, the second derivative can be calculated 
analytically (rather approximate) and finding edges is done by imposing a 
condition over the model parameters. \\

@@

In conjunction with this paper we present a detailed and well commented source code, which 
implements the described algorithms. Also in the appendices (sections \ref{sec:appendix1} 
and \ref{sec:appendix2}) we show some common mathematical developments, and a commented 
explanation of the source code. This work also provides tools for code documentation 
(see section \ref{sec:appendix2}). \\

\nocite{IPOL}
%-------------------------------------------------------------------------------
\section{Algorithms based on first derivative}
\label{sec:first}

The usual tool to find the amplitude and direction of changes in 
intensity of an image $f$ is the gradient operator (denoted as $\nabla$), defined 
as the vector:
\begin{equation}
	\nabla f = 
				\begin{bmatrix} 
					g_x \\ g_y
				\end{bmatrix}
	=				
				\begin{bmatrix} 
					\cfrac{\partial f}{\partial x} \\ \cfrac{\partial f}{\partial y} \\
				\end{bmatrix}
\end{equation}

The magnitude ($M$) and direction ($\alpha$) of the gradient vector $\nabla f$ at location $(x,y)$
are calculated as:
\begin{equation}
	\begin{cases}
		M(x,y) = \sqrt{g_x^2 + g_y^2} \\
		\alpha(x,y) = \tan^{-1} \begin{bmatrix} g_x \\ g_y \end{bmatrix} \\
	\end{cases}
\end{equation}

@@

We need to compute partial derivatives $\partial f/\partial x$ and $\partial f/\partial y$ 
at every pixel in the image. Since we are dealing with digital images, numerical approximations 
of the partial derivatives over a neighborhood about a point are required. In the following we will study 
the methods of Roberts, Prewitt and Sobel, whose main difference is how they perform this calculation.\\

@@

The most usual approach of the first derivative is to take the Taylor expansion of first order with
a small $h$. Approximating the first derivative:
\begin{equation}
	f'(x) \simeq \frac{f(x+h) - f(x)}{h}\\
\end{equation}
The image has discrete points of coordinates $(i,j)$, so taking $x=i*h$ and $y=j*h$, we approximate the 
first derivative of the image based on the intensity values ​​at points of the image as:
\begin{equation}
\label{eq:roberts1}
	g_x = \frac{\partial f(x,y)}{\partial x} = f(i+1,j) - f(i,j)
\end{equation}
and
\begin{equation}
\label{eq:roberts2}
	g_y = \frac{\partial f(x,y)}{\partial y} = f(i,j+1) - f(i,j)\\
\end{equation}

Equations \ref{eq:roberts1} and \ref{eq:roberts2} can be implemented for all values of $x$ and $y$
by filtering the image $f(x,y)$ with the 1-D masks in figure \ref{fig:1dmasks}. But when we want 
to find diagonal edges, we need 2-D masks. The \textit{Roberts operators} are one of the earliest 
attempts to use 2-D masks for this purpose. These operators are based on implementing the diagonal 
diferences implemented by filtering an image with the masks in figure \ref{fig:roberts}.\\

@@

@@

A slight variation of the Prewitt operators use more weight on the central coefficients of the 
difference. It can be shown that using a $2$ in the center location provides image smoothing. This
variation is implemented using masks in figure \ref{fig:sobel}. These operators are called 
\textit{Sobel operators}.\\

@@

The Prewitt masks are simpler to implement than the Sobel masks, but the computational difference
between them is not an issue. It is preferable to use Sobel operators, because they have better
noise-suppression characteristics (smoothing). Noise suppression is an important issue when dealing
with derivatives.\\

%-------------------------------------------------------------------------------
\section{Algorithms based on second derivative}
\label{sec:second}

@@

The two algorithms that we will see in this section (Marr-Hildreth \cite{AIM-518} [1980] and Haralick \cite{bb20239} [1987]) 
are based on the second derivative of the image, and both take steps to reduce noise before 
detecting edges in the image.\\

@@

The Marr-Hildreth algorithm is a method of detecting edges in digital 
images. It's based on find zero crossing points of the second derivative
of the image. This can be done in several ways. In this work we implemented 
two different ways; convolving the image with a Gaussian kernel and then 
approximating the second derivative (Laplacian) with a 3x3 kernel, or 
convolving the image with a kernel calculated as the Laplacian of a 
Gaussian function. There are more ways to do so, for example, using 
recursive Gaussian filters \cite{Deriche1993Recursively}.\\

The algorithm is divided in three steps, each one described later:
\begin{enumerate}
	\item Grayscale conversion of the input image (see section \ref{sec:appendix1}).
	\item Convolution of the image with:
	\begin{itemize}
		\item a Laplacian of Gaussian (LoG) kernel. (or)
		\item a Gaussian kernel and then a Laplacian operator.
	\end{itemize}
	\item Search of zero crossing points in the filtered image.\\
\end{enumerate}

Some auxiliary functions were implemented for computing the operations
needed in the algorithm, such as Gaussian kernel and Laplacian of a Gaussian 
kernel generation, and 2-D convolution of an image with a given kernel, 
with zero padding.\\

%-------------------------------------------------------------------------------
\subsubsection{Gaussian and LoG kernels}

The Marr-Hildreth algorithm consists of convolving the LoG filter with an input image, $f(x,y)$.
\begin{equation}\label{eq:log}
  g(x,y) = [\nabla^2G(x,y)]\star f(x,y)
\end{equation}
and then finding the zero crossings of $g(x,y)$ to determine the location of edges in $f(x,y)$. 
Because these are linear processes, equation \ref{eq:log} can be written also as
\begin{equation}
  g(x,y) = \nabla^2[G(x,y)\star f(x,y)]
\end{equation}
indicating that we can smooth the image first with a Gaussian filter and then compute the Laplacian of the result.\\

Then we need to generate both Gaussian kernel and LoG kernel (see section \ref{sec:appendix2}).\\

The Marr-Hildreth edge-detection algorithm may be summarized as follows:
\begin{enumerate}
	\item Filter the input image with a $n \times n$ Gaussian lowpass filter obtained by sampling equation \ref{eq:gaussian_function}.
	\item Compute the Laplacian of the image resulting from step 1, using, for example, the $3\times3$ mask:
			$\begin{bmatrix}
			1 &  1 & 1 \\
			1 & -8 & 1 \\
			1 &  1 & 1 \\
			\end{bmatrix}$\footnote{Steps 1 and 2 can be made into one, using a $n\times n$ LoG lowpass filter obtained by sampling equation \ref{eq:log_function}.}
	\item Find the zero crossings of the image from step 2.
\end{enumerate}

%-------------------------------------------------------------------------------
\subsubsection{Zero crossing}

A zero crossing at pixel $p$ implies that the signs of at least two opposite neighboring pixels are 
different. There are four cases to test: left/right, up/down, and the two diagonals. In this case 
we work with a threshold, so that not only the signs of the opposite pixels must differ, also their 
difference in absolute value must be greater than a certain threshold.\\

Zero crossing detection is the key feature of the Marr-Hildreth edge detection method. The technique 
presented in the previous paragraph is attractive for its simplicity of implementation and its low 
computational cost. In general it is a technique that yields good results, but if more precision in 
finding the zero crossings is needed, methods for finding zero crossings with subpixel accuracy 
should be used.\\

%-------------------------------------------------------------------------------
\subsection{The \textit{Haralick} algorithm}

The idea behind the Haralick edge detector is identical to that of the previous method; find zeros in 
the second derivative of the image. In this method, we smooth the input image through local bi-cubic
polynomial fitting. Then, calculating the second derivative analytically, we can find an expression 
equivalent to cancel the second derivative of the polynomial as a function of the parameters of it.\footnote{This implementation is slightly 
different to the traditional implementation of the Haralick algorithm. We do not use the condition concerning the third derivative (to be negative), in order to implement two edge detection methods using only the second derivative (along the Marr-Hildreth algorithm).}\\

@@

We adjust the surrounding neighborhood of a point $(x,y)$ in the image $f$ using the following bi-cubic polynomial:
\begin{equation}
	\label{eq:bicubic}
	f(x,y) = k_1 + k_2x + k_3y + k_4x^2 + k_5xy + k_6y^2 + k_7x^3 + k_8x^2y + k_9xy^2 + k_{10}y^3\\
\end{equation}

To solve this problem, we need to take more neighbors than coefficients to be adjusted. The smallest 
neighborhood of odd dimensions has size $5\times5$. Now it is possible to convolve the input image
with some precomputed masks, instead of finding the least square solution to find the coefficients 
$k_1\dots k_{10}$ at each point $(x,y)$.\\

Consider 25 points in a small neighborhood of a point $(x,y)$ in the image. We use a least square
formulation to obtain the coefficients (25 points give equal number of equations):

\begin{equation*}
	\begin{array}{l}
		f_1 = f(x_1,y_1) = k_1 + k_2x_1 + k_3y_1 + k_4x_1^2 + k_5x_1y_1 + k_6y_1^2 + k_7x_1^3 + k_8x_1^2y_1 + k_9x_1y_1^2 + k_{10}y_1^3\\
		f_2 = f(x_2,y_2) = k_1 + k_2x_2 + k_3y_2 + k_4x_2^2 + k_5x_2y_2 + k_6y_2^2 + k_7x_2^3 + k_8x_2^2y_2 + k_9x_2y_2^2 + k_{10}y_2^3\\
		\vdots \\
		f_{25} = f(x_{25},y_{25}) = k_1 + k_2x_{25} + k_3y_{25} + k_4x_{25}^2 + k_5x_{25}y_{25} + k_6y_{25}^2 + k_7x_{25}^3 + k_8x_{25}^2y_{25} + k_9x_{25}y_{25}^2 + k_{10}y_{25}^3\\
	\end{array}
\end{equation*}

Using matrix notation:

\begin{equation*}
	\begin{bmatrix} 
		f_1		\\ 
		f_2		\\ 
		\vdots	\\
		f_{25}
	\end{bmatrix} 
	= 
	\begin{bmatrix} 
		1 		& x_1 		& y_1 		& x_1^2 	& x_1y_1 		& y_1^2 	& \hdots 	& y_1^3 	\\
		1 		& x_2 		& y_2 		& x_2^2 	& x_2y_2 		& y_2^2 	& \hdots 	& y_2^3 	\\
		\vdots	& \vdots	& \vdots	& \vdots	& \vdots		& \vdots	& \ddots	& \vdots	\\
		1 		& x_{25}	& y_{25}	& x_{25}^2 	& x_{25}y_{25} 	& y_{25}^2 	& \hdots 	& y_{25}^3
	\end{bmatrix}
	\times
	\begin{bmatrix}
		k_1		\\
		k_2		\\
		\vdots	\\
		k_{10}
	\end{bmatrix}
	\Rightarrow \mathbf{f} = \mathbf{A}\mathbf{k}\\
\end{equation*}

Then:

\begin{equation*}
	(\mathbf{A}^T\mathbf{A})^{-1}\mathbf{A}^T\mathbf{f} = \mathbf{k} \ \ \Rightarrow \ \ \mathbf{k} = \mathbf{B}\mathbf{f}\\
\end{equation*}

$\mathbf{B}$ is a $10\times25$ matrix:

\begin{equation*}
	\begin{bmatrix}
		b_{1,1}		& b_{1,2}	& \hdots	& b_{1,25}	\\
		b_{2,1}		& b_{2,2}	& \hdots	& b_{2,25}	\\
		\vdots		& \vdots	& \ddots	& \vdots	\\
		b_{10,1}	& b_{10,2}	& \hdots	& b_{10,25}
	\end{bmatrix}
	\times
	\begin{bmatrix}
		f_1		\\
		f_2		\\
		\vdots	\\
		f_{25}
	\end{bmatrix}
	=
	\begin{bmatrix}
		k_1		\\
		k_2		\\
		\vdots	\\
		k_{10}
	\end{bmatrix}\\
\end{equation*}

For each coefficient ($i$ from $1$ to $10$):

\begin{equation}
	\label{eq:coefficients}
	k_i = b_{i,1}f_1 + b_{i,2}f_2 + b_{i,3}f_3 + \hdots + b_{i,25}f_{25} \ \ \Rightarrow \ \ \mathbf{k_i} = \mathbf{f}\star\mathbf{b_i}\\
\end{equation}

where:

>>>>

%-------------------------------------------------------------------------------
% Edge detectors review v6
% by Haldo Spontón & Juan Cardelino
% IPOL 2012
%-------------------------------------------------------------------------------

\documentclass{ipol}
\ipolSetYear{2012}
\ipolSetDOI{10.XXXX/ipol.2012.sc-arced}

\usepackage{hyperref,verbatim,graphicx,amsmath,amssymb,amssymb,dsfont}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{sidecap,subfig,array}
\usepackage[table]{xcolor}
\usepackage{multicol}
\usepackage{listings}
\usepackage{color}
%\usepackage{algorithm}
\usepackage{algorithmic}
\newtheorem{theorem}{Theorem}
\numberwithin{equation}{section}
\numberwithin{table}{section}
\numberwithin{figure}{section}
\begin{document}
\bibliographystyle{ieeetr}
\lstset{
	backgroundcolor=\color[rgb]{0.94,0.94,0.94},
    tabsize=4,
    inputencoding=utf8x,
         extendedchars=\true,
         language=C,
         basicstyle=\scriptsize,
                 showtabs=false,
        showspaces=false,
        showstringspaces=false,
        identifierstyle=\ttfamily,
        keywordstyle=\color[rgb]{0,0,1},
        commentstyle=\color[rgb]{0.133,0.545,0.133},
        stringstyle=\color[rgb]{0.627,0.126,0.941},
}

%-------------------------------------------------------------------------------

\title{A Review of Classic Edge Detectors}
\author{Haldo Spont\'on,\\
        Juan Cardelino}
\date{}
\ipolMaketitle

%-------------------------------------------------------------------------------

\begin{ipolAbstract}
In this paper some of the classic alternatives for edge detection in digital images are studied. The main idea 
behind edge detection is to find where there are abrupt changes in the intensity of an image. 
As a first approach, the first derivative is used to find the changes of intensity in several algorithms 
such as Sobel, Prewitt and Roberts. The second derivative is also used, for example in algorithms 
like Marr-Hildreth and Haralick.\\
Results obtained from a qualitative point of view (perceptual) and from a quantitative 
point of view (number of operations, execution time) are compared, considering different forms to convolve an 
image with a kernel (step required in some of the algorithms).
\end{ipolAbstract}

%-------------------------------------------------------------------------------

\begin{ipolCode}
The C implementation of the algorithms implemented in this work (beta version) can be
downloaded from:
\begin{center}
\href{http://iie.fing.edu.uy/~haldos/downloads/edge_detectors_v0.1.tar.gz}{\url{http://iie.fing.edu.uy/~haldos/downloads/edge_detectors_v0.1.tar.gz}}.
\end{center}
You can also follow this code project in this GIT repository: 
\begin{center}
\url{git://github.com/haldos/edges.git}
\end{center}
\end{ipolCode}

%-------------------------------------------------------------------------------

%\begin{ipolSupp}
%\end{ipolSupp}

%-------------------------------------------------------------------------------

\section{Introduction}
\label{sec:intro}

@@

In the 70's, edge detection methods were based on using small operators 
(such as Sobel masks), attempting to compute an approximation of the
first derivative of the image \cite{Gonzalez2007Digital}. In section ~\ref{sec:first} the 
behavior of such algorithms is described, as an introduction to a more sophisticated 
analysis into the edge detection process.\\

In 1980 Marr and Hildreth \cite{AIM-518} argued that intensity changes are not independent 
of image scale, so edge detection requires the use of different size 
operators. They also argued that a sudden intensity change will be seen 
as a peak (or trough) in the first derivative or, equivalently, as a zero 
crossing in the second derivative. The bulk of this work focuses on the 
study of this algorithm, presented in Section ~\ref{sec:second}. In this section 
the Haralick's algorithm \cite{bb20239} is also presented, which is based on second derivative too.
This algorithm has the particularity of proposing a model to locally approximate 
the image around a point. Then, with this model, second derivative can be calculated 
analytically (rather approximate) and finding edges is done by imposing a 
condition over the model parameters. \\

@@

Within this paper, a detailed and well commented source code is presented, which 
implements the described algorithms. In sections \ref{sec:appendix1} 
and \ref{sec:appendix2} some common mathematical developments, and a commented 
explanation of the source code are shown. This work also provides tools for code documentation 
(see section \ref{sec:appendix2}).

\nocite{IPOL}

%-------------------------------------------------------------------------------

\section{Algorithms based on first derivative}
\label{sec:first}

The usual tool to find the amplitude and direction of changes in 
intensity of an image $f$ is the gradient operator (denoted as $\nabla$), defined 
as the vector
\begin{equation}
	\nabla f = 
				\begin{bmatrix} 
					g_x \\ g_y
				\end{bmatrix}
	=				
				\begin{bmatrix} 
					\cfrac{\partial f}{\partial x} \\ \cfrac{\partial f}{\partial y} \\
				\end{bmatrix}.
\end{equation}

The magnitude ($M$) and direction ($\alpha$) of the gradient vector $\nabla f$ at location $(x,y)$
are calculated as
\begin{equation}
\label{eq:mag_alpha}
	\begin{cases}
		M(x,y) = \sqrt{g_x^2 + g_y^2} \\
		\alpha(x,y) = \tan^{-1} \begin{bmatrix} g_x \\ g_y \end{bmatrix} .\\
	\end{cases}
\end{equation}

@@

The computation of partial derivatives $\partial f/\partial x$ and $\partial f/\partial y$ 
at every pixel in the image is required. Dealing with digital images, numerical approximations 
of the partial derivatives over a neighborhood about a point are required. In the following,
the methods of Roberts, Prewitt and Sobel will be studied, whose main difference is how they perform this calculation.

%-------------------------------------------------------------------------------

@@

The most usual approach of the first derivative is to take the Taylor expansion of first order with
a small $h$. Approximating the first derivative
\begin{equation}
	f'(x) \simeq \frac{f(x+h) - f(x)}{h}.\\
\end{equation}
The image has discrete points of coordinates $(i,j)$, so taking $x=i*h$ and $y=j*h$, the 
first derivative of the image is approximated, based on the intensity values ​​at points of the image, as:
\begin{equation}
\label{eq:roberts1}
	g_x = \frac{\partial f(x,y)}{\partial x} = f(i+1,j) - f(i,j)
\end{equation}
and
\begin{equation}
\label{eq:roberts2}
	g_y = \frac{\partial f(x,y)}{\partial y} = f(i,j+1) - f(i,j).\\
\end{equation}

Equations \ref{eq:roberts1} and \ref{eq:roberts2} can be implemented for all values of $x$ and $y$
by filtering the image $f(x,y)$ with the 1-D masks in figure \ref{fig:1dmasks}. But when finding 
diagonal edges is desired, 2-D masks are needed. The \textit{Roberts operators} are one of the earliest 
attempts to use 2-D masks for this purpose. These operators are based on implementing the diagonal 
diferences implemented by filtering an image with the masks in figure \ref{fig:roberts}. By convention 
in these figures, x-coordinates grow from left to right and y-coordinates grow top-down. \\

@@

%-------------------------------------------------------------------------------

@@

A slight variation of the Prewitt operators uses more weight on the central coefficients of the 
difference. It can be shown that using a $2$ in the center location provides image smoothing. This
variation is implemented using masks in figure \ref{fig:sobel}. These operators are called 
\textit{Sobel operators}.\\

@@

The Prewitt masks are simpler to implement than the Sobel masks, but the computational difference
between them is not an issue. It is preferable to use Sobel operators, because they are better localized, 
and their filter are also less aliased, because of the Gaussian shape of $(1\ 2\ 1)$.

% they have better noise-suppression characteristics (smoothing). Noise suppression is an important issue when dealing with derivatives.

%-------------------------------------------------------------------------------

\subsection{Edges image}

Once the respective operators are applied, two images per method are obtained, containing the approximation of horizontal
and vertical components of the first derivative, respectively. Edges image is obtained by applying a 
threshold to the first derivative magnitude image, calculated using the equation \ref{eq:mag_alpha}. 
A black and white image is obtained, in which edge points are indicated in white.

%-------------------------------------------------------------------------------

\subsection{Pseudo-code}

Pseudo-code of implemented algorithms is shown in algorithm \ref{algo:fded}. \\

\begin{algorithm}
\caption{First derivative edge detection algorithms.}
\label{algo:fded}
\begin{algorithmic}[1]
\REQUIRE input image, threshold $th$.
\STATE $im_{ORIG} \leftarrow$ input image
\IF{$im_{ORIG}$ is grayscale}
	\STATE $im \leftarrow im_{ORIG}$
\ELSE 
	\STATE $im \leftarrow (6968im_{ORIG,R}+23434im_{ORIG,G}+2366im_{ORIG,B})/32768$ \COMMENT{Grayscale conversion, assuming RGB image.}
\ENDIF
\STATE Define $operator_x$ and $operator_y$ \COMMENT{Roberts, Prewitt or Sobel.}
\STATE $g_x \leftarrow$ convolution($im$,$operator_x$)
\STATE $g_y \leftarrow$ convolution($im$,$operator_y$)
\STATE $max_M \leftarrow 0$
\FORALL{pixel $i$ in image}
	\STATE $M[i] \leftarrow \sqrt{g_x^2+g_y^2}$
	\IF{$M[i]>max_M$}
		\STATE $max_M \leftarrow M[i]$
	\ENDIF
\ENDFOR
\FORALL{pixel $i$ in image}
	\IF{$M[i] \geq th \times max_M$}
		\STATE $im_{OUT}[i] \leftarrow 255$
	\ELSE
		\STATE $im_{OUT}[i] \leftarrow 0$
	\ENDIF
\ENDFOR
\RETURN output image $\leftarrow im_{OUT}$
\end{algorithmic}
\end{algorithm}

%-------------------------------------------------------------------------------

\section{Algorithms based on second derivative}
\label{sec:second}

@@

The two algorithms that are going to be shown in this section (Marr-Hildreth \cite{AIM-518} [1980] and Haralick \cite{bb20239} [1987]) 
are based on the second derivative of the image, and both take steps to reduce noise before 
detecting edges in the image.

%-------------------------------------------------------------------------------

@@

The Marr-Hildreth algorithm is a method of detecting edges in digital 
images. It is based on find zero crossing points of the second derivative
of the image. This can be done in several ways. Two different ways of doing 
this are implemented implemented in this work; convolving the image with a Gaussian kernel and then 
approximating the second derivative (Laplacian) with a 3x3 kernel, or 
convolving the image with a kernel calculated as the Laplacian of a 
Gaussian function. There are more ways to do so, for example, using 
recursive Gaussian filters \cite{Deriche1993Recursively}.\\

The algorithm is divided in three steps, each one described later:
\begin{enumerate}
	\item Grayscale conversion of the input image (see section \ref{sec:appendix1}).
	\item Convolution of the image with:
	\begin{itemize}
		\item a Laplacian of Gaussian (LoG) kernel. (or)
		\item a Gaussian kernel and then a Laplacian operator.
	\end{itemize}
	\item Search of zero crossing points in the filtered image.
\end{enumerate}

Some auxiliary functions were implemented for computing the operations
needed in the algorithm, such as Gaussian kernel and Laplacian of a Gaussian 
kernel generation, and 2-D convolution of an image with a given kernel, 
with zero padding.

%-------------------------------------------------------------------------------

\subsubsection{Gaussian and LoG kernels}

The Marr-Hildreth algorithm consists of convolving the input image $f(x,y)$ with a LoG filter;
\begin{equation}\label{eq:log}
  g(x,y) = [\nabla^2G(x,y)]\star f(x,y), 
\end{equation}
and then finding the zero crossings of $g(x,y)$ to determine the location of edges in $f(x,y)$. 
Because these are linear processes, equation \ref{eq:log} can be written also as
\begin{equation}
  g(x,y) = \nabla^2[G(x,y)\star f(x,y)]
\end{equation}
indicating that image can be smoothed with a Gaussian filter first, and then compute the Laplacian of the result.\\

Then, generation of both Gaussian and LoG kernel is required (see section \ref{sec:appendix2}).\\

Marr-Hildreth edge-detection algorithm may be summarized as follows:
\begin{enumerate}
	\item Filter the input image with a $n \times n$ Gaussian lowpass filter obtained by sampling equation \ref{eq:gaussian_function}.
	\item Compute the Laplacian of the image resulting from step 1, using, for example, the $3\times3$ mask:
			$\begin{bmatrix}
			1 &  1 & 1 \\
			1 & -8 & 1 \\
			1 &  1 & 1 \\
			\end{bmatrix}$\footnote{Steps 1 and 2 can be made into one, using a $n\times n$ LoG lowpass filter obtained by sampling equation \ref{eq:log_function}.}
	\item Find the zero crossings of the image from step 2.
\end{enumerate}

%-------------------------------------------------------------------------------

\subsubsection{Zero crossing}

A zero crossing at pixel $p$ implies that the signs of at least two opposite neighboring pixels are 
different. There are four cases to test: left/right, up/down, and the two diagonals. In this case 
a threshold is used, so that not only the signs of the opposite pixels must differ, also their 
difference in absolute value must be greater than a certain threshold.\\

Zero crossing detection is the key feature of the Marr-Hildreth edge detection method. The technique 
presented in the previous paragraph is attractive for its simplicity of implementation and its low 
computational cost. In general it is a technique that yields good results, but if more precision in 
finding the zero crossings is needed, methods for finding zero crossings with subpixel accuracy 
should be used.

%-------------------------------------------------------------------------------

\subsubsection{Pseudo-code}

Pseudo-code of Marr-Hildreth's algorithm is shown in algorithm \ref{algo:mh}.

\begin{algorithm}[t]
\caption{Marr-Hildreth edge detection algorithm.}
\label{algo:mh}
\begin{algorithmic}[1]
\REQUIRE input image, standard deviation $\sigma$, kernel size $n$ and zero-crossing threshold $tzc$.
\STATE $im_{ORIG} \leftarrow$ input image
\IF{$im_{ORIG}$ is grayscale}
	\STATE $im \leftarrow im_{ORIG}$
\ELSE 
	\STATE $im \leftarrow (6968im_{ORIG,R}+23434im_{ORIG,G}+2366im_{ORIG,B})/32768$ \COMMENT{Grayscale conversion, assuming RGB image.}
\ENDIF
\STATE $kernel \leftarrow$ Generated Gaussian or LoG kernel
\STATE $im_{SMOOTHED} \leftarrow$ convolution($im$,$kernel$)
\IF{Gaussian kernel}
	\STATE Define Laplacian approx. operator $laplacian$
	\STATE $im_{LAPL} \leftarrow$ convolution($im$,$laplacian$)
\ELSE
	\STATE $im_{LAPL} \leftarrow im_{SMOOTHED}$
\ENDIF
\STATE $max_L \leftarrow 0$
\FORALL{pixel $i$ in image $im_{LAPL}$}
	\IF{$im_{LAPL}[i]>max_L$}
		\STATE $max_L \leftarrow im_{LAPL}[i]$
	\ENDIF
\ENDFOR
\FORALL{pixel $i$ in image $im_{LAPL}$, except borders}
	\IF{opposite neighbors of pixel $i$ have opposite signs in $im_{LAPL}$}
		\STATE $im_{OUT}[i] \leftarrow 255$
	\ELSE
		\STATE $im_{OUT}[i] \leftarrow 0$
	\ENDIF
\ENDFOR
\RETURN output image $\leftarrow im_{OUT}$
\end{algorithmic}
\end{algorithm}

%-------------------------------------------------------------------------------

\subsection{The \textit{Haralick} algorithm}

The idea behind the Haralick edge detector is identical to that of the previous method; find zeros in 
the second derivative of the image. In this method, the input image is smoothed through local bi-cubic
polynomial fitting. Then, calculating the second derivative analytically, it is possible to find an equivalent expression to cancel the second derivative of the polynomial as a function of its parameters.\footnote{This implementation is slightly 
different from the traditional implementation of the Haralick algorithm. We do not use the condition concerning the third derivative (to be negative), in order to implement two edge detection methods using only the second derivative (along the Marr-Hildreth algorithm).}

%-------------------------------------------------------------------------------

@@

The surrounding neighborhood of a point $(x,y)$ in the image $f$ is adjusted using the following bi-cubic polynomial
\begin{equation}
	\label{eq:bicubic}
	f(x,y) = k_1 + k_2x + k_3y + k_4x^2 + k_5xy + k_6y^2 + k_7x^3 + k_8x^2y + k_9xy^2 + k_{10}y^3.\\
\end{equation}

To solve this problem, it is necessary to take more neighbors than coefficients to be adjusted. The smallest 
neighborhood of odd dimensions has size $5\times5$. Now it is possible to convolve the input image
with some precomputed masks, instead of finding the least square solution to find the coefficients 
$k_1\dots k_{10}$ at each point $(x,y)$.\\

Consider 25 points in a small neighborhood of a point $(x,y)$ in the image. A least square
formulation is used to obtain the coefficients (25 points give equal number of equations),

\begin{equation*}
	\begin{array}{l}
		f_1 = f(x_1,y_1) = k_1 + k_2x_1 + k_3y_1 + k_4x_1^2 + k_5x_1y_1 + k_6y_1^2 + k_7x_1^3 + k_8x_1^2y_1 + k_9x_1y_1^2 + k_{10}y_1^3\\
		f_2 = f(x_2,y_2) = k_1 + k_2x_2 + k_3y_2 + k_4x_2^2 + k_5x_2y_2 + k_6y_2^2 + k_7x_2^3 + k_8x_2^2y_2 + k_9x_2y_2^2 + k_{10}y_2^3\\
		\vdots \\
		f_{25} = f(x_{25},y_{25}) = k_1 + k_2x_{25} + k_3y_{25} + k_4x_{25}^2 + k_5x_{25}y_{25} + k_6y_{25}^2 + k_7x_{25}^3 + k_8x_{25}^2y_{25} + k_9x_{25}y_{25}^2 + k_{10}y_{25}^3 .\\
	\end{array}
\end{equation*}

Using matrix notation,

\begin{equation*}
	\begin{bmatrix} 
		f_1		\\ 
		f_2		\\ 
		\vdots	\\
		f_{25}
	\end{bmatrix} 
	= 
	\begin{bmatrix} 
		1 		& x_1 		& y_1 		& x_1^2 	& x_1y_1 		& y_1^2 	& \hdots 	& y_1^3 	\\
		1 		& x_2 		& y_2 		& x_2^2 	& x_2y_2 		& y_2^2 	& \hdots 	& y_2^3 	\\
		\vdots	& \vdots	& \vdots	& \vdots	& \vdots		& \vdots	& \ddots	& \vdots	\\
		1 		& x_{25}	& y_{25}	& x_{25}^2 	& x_{25}y_{25} 	& y_{25}^2 	& \hdots 	& y_{25}^3
	\end{bmatrix}
	\times
	\begin{bmatrix}
		k_1		\\
		k_2		\\
		\vdots	\\
		k_{10}
	\end{bmatrix}
	\Rightarrow \mathbf{f} = \mathbf{A}\mathbf{k} .\\
\end{equation*}

Then,

\begin{equation*}
	(\mathbf{A}^T\mathbf{A})^{-1}\mathbf{A}^T\mathbf{f} = \mathbf{k} \ \ \Rightarrow \ \ \mathbf{k} = \mathbf{B}\mathbf{f} .\\
\end{equation*}

$\mathbf{B}$ is a $10\times25$ matrix,

\begin{equation*}
	\begin{bmatrix}
		b_{1,1}		& b_{1,2}	& \hdots	& b_{1,25}	\\
		b_{2,1}		& b_{2,2}	& \hdots	& b_{2,25}	\\
		\vdots		& \vdots	& \ddots	& \vdots	\\
		b_{10,1}	& b_{10,2}	& \hdots	& b_{10,25}
	\end{bmatrix}
	\times
	\begin{bmatrix}
		f_1		\\
		f_2		\\
		\vdots	\\
		f_{25}
	\end{bmatrix}
	=
	\begin{bmatrix}
		k_1		\\
		k_2		\\
		\vdots	\\
		k_{10}
	\end{bmatrix} .\\
\end{equation*}

For each coefficient ($i$ from $1$ to $10$),

\begin{equation}
	\label{eq:coefficients}
	k_i = b_{i,1}f_1 + b_{i,2}f_2 + b_{i,3}f_3 + \hdots + b_{i,25}f_{25} \ \ \Rightarrow \ \ \mathbf{k_i} = \mathbf{f}\star\mathbf{b_i},\\
\end{equation}

where
